---
title: "SBC Data Exploration"
author: "Mary Fisher"
date: "February 28, 2020"
output: 
  html_document:
    toc: yes
    toc_float: yes
---

A quick data exploration for correlation between variables.

Research Questions:

1. How does fish / invertebrate community abundance and diversity vary according to giant kelp size and abundance?

2. How are oceanographic conditions and herbivore abundance correlated with giant kelp size and abundance?

This script was rerun on Mar 5, 2020, for observations at the site (instead of transect) level.

<br>

```{r setup, include=FALSE}
rm(list=ls())

knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(janitor)
library(here)
library(vegan)
library(lubridate)
library(corrplot)
library(MASS)
```



## Data 

For question 1, I need the kelp, fish, and invertebrate data.
```{r}
kelp <- read.csv(here::here("data","Annual_Kelp_MeanSize_MeanAbund_2001-2018.csv"))
fish <- read.csv(here::here("data","Annual_Fish_RichnessAbundDiversity_2001-2018.csv")) %>% 
  filter(survey=="FISH") %>% dplyr::select(-date, -survey)
invert <- read.csv(here::here("data","Annual_Invert_RichnessAbundDiversity_2001-2018.csv")) %>% dplyr::select(-date, -franciscanus.abundance,-purpuratus.abundance,-urchin.abundance)

# aggregate fish / invert by site
fish <- fish %>%
  pivot_longer(cols=colnames(fish)[5:13]) %>%
  group_by(year,month,site,name) %>%
  summarise(fish.mean=mean(value,na.rm=TRUE)) %>%
  pivot_wider(names_from=name,values_from=fish.mean)
invert <- invert %>%
  pivot_longer(cols=colnames(invert)[6:14]) %>%
  group_by(year,month,site,survey,name) %>%
  summarise(fish.mean=mean(value,na.rm=TRUE)) %>%
  pivot_wider(names_from=c(survey,name),values_from=fish.mean)

q1dat_fish <- left_join(kelp,fish, by=c("year","month","site"))
q1dat_invert <- left_join(kelp,invert, by=c("year","month","site"))
colnames(q1dat_fish)
```
<br>

For question 2, I need the kelp, oceanographic, and urchin abundance data.
```{r}
urchins <- read.csv(here::here("data","Annual_Invert_RichnessAbundDiversity_2001-2018.csv")) %>% 
  dplyr::select(year,month,site,transect, franciscanus.abundance,purpuratus.abundance,urchin.abundance) %>%
  pivot_longer(cols=c(franciscanus.abundance,purpuratus.abundance,urchin.abundance)) %>%
  group_by(year,month,site,name) %>%
  summarise(urchin.means=mean(value,na.rm=TRUE)) %>%
  pivot_wider(names_from=name,values_from=urchin.means)
waterchem <- read.csv(here::here("data","Annual_CTD_WaterChemistry_MeanMax_2000-2018.csv"))
temp <- read.csv(here::here("data","Seasonal_Mean_Bottom_Temp_2002-2019.csv")) %>%
  dplyr::select(year,site,season,mean.temp,var.temp) %>%
  rename(mean_temp=mean.temp,var_temp=var.temp) %>%
  pivot_longer(cols=c(mean_temp,var_temp)) %>%
  pivot_wider(names_from=c(season,name), values_from=value)

dat2_urchins <- left_join(kelp,urchins,by=c("year","month","site"))
dat2_water <- left_join(kelp,waterchem,by=c("site"="station","year")) %>%
  left_join(temp,by=c("site","year"))
head(dat2_urchins); head(dat2_water)
```
<br>




## Question 1

### Fish Community Composition
How does fish community abundance and diversity vary according to giant kelp abundance?

```{r}
abund_fish <- q1dat_fish %>% dplyr::select(-mapy_mean_fronds,-mapy_max_fronds,-mapy_mean_hld,-mapy_max_hld)
corrplot(cor(abund_fish[,4:14],use="pairwise.complete.obs"), type="upper")
```
<br>

How does fish community abundance and diversity vary according to giant kelp size?

```{r}
size_fish <- q1dat_fish %>% dplyr::select(-mapy_mean_count,-mapy_mean_density)
corrplot(cor(size_fish[,4:16],use="complete.obs"), type="upper")
```
<br>

### Invertebrate Community Composition
How does invertebrate community abundance and diversity vary according to giant kelp abundance?

```{r}
abund_invert <- q1dat_invert %>% dplyr::select(-mapy_mean_fronds,-mapy_max_fronds,-mapy_mean_hld,-mapy_max_hld)
corrplot(cor(abund_invert[,4:14],use="complete.obs"), type="upper")
corrplot(cor(abund_invert[,c(4:5,15:22)],use="complete.obs"), type="upper")
```
<br>

How does invertebrate community abundance and diversity vary according to giant kelp size?

```{r}
size_invert <- q1dat_invert %>% dplyr::select(-mapy_mean_density,-mapy_mean_count)
corrplot(cor(abund_invert[,4:14],use="complete.obs"), type="upper")
corrplot(cor(abund_invert[,c(4:5,15:22)],use="complete.obs"), type="upper")
```
<br>

### Fish v. Invertebrate Communities
```{r warning=FALSE}
invert_fish <- left_join(invert,fish,by=c("year","month","site"))

corrplot(cor(invert_fish[,c(4:12)],invert_fish[,c(22:30)],use="complete.obs"), type="upper")
corrplot(cor(invert_fish[,c(13:21)],invert_fish[,c(22:30)],use="complete.obs"), type="upper")
```



## Question 2

How does kelp size / abundance vary according to sea urchin density?

```{r}
corrplot(cor(dat2_urchins[,4:12],use="complete.obs"), type="upper")
```
<br>

How does kelp size / abundance vary according to water chemistry?

```{r}
kelp_chem_means <- filter(dat2_water, site %in% unique(waterchem$station)) %>% dplyr::select(4:9,16:21)
kelp_chem_max <- filter(dat2_water, site %in% unique(waterchem$station)) %>% dplyr::select(4:15)
corrplot(cor(kelp_chem_max,use="complete.obs"), type="upper")
corrplot(cor(kelp_chem_means,use="complete.obs"), type="upper")
```
<br>

How does kelp size / abundance vary according to temperature?
```{r}
kelp_temp <- filter(dat2_water, !is.na(summer_mean_temp))

corrplot(cor(kelp_temp[,c(4:9,22:25)],use="complete.obs"), type="upper")
```






