---
title: "Transformation of Response Variables"
author: "Mary Fisher"
date: "March 6, 2020"
output: 
  html_document:
    toc: yes
    toc_float: yes
  
---


Check the distributions for each predictor variable, and use the box-cox function on a simple linear model to determine the suggested transformation.

**NOTE:** most of the non-normal response variables follow a poisson distribution, or a zero-inflated poisson distribution. instead of applying a transformation to the response *and* predictors, we could also just use a GLM with a Poisson link function.
<br>

```{r setup, include=FALSE}
rm(list=ls())

knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(janitor)
library(here)
library(vegan)
library(lubridate)
library(corrplot)
library(MASS)

theme_set(theme_bw())
```
<br>

Our model data:
```{r echo=TRUE}
mydat1 <- read.csv(here::here("data","kelp_prediction_data_complete.csv")) %>%
  clean_names()
colnames(mydat1)

mydat2 <- read.csv(here::here("data","biomass_prediction_data_waterchem_bysite_complete.csv")) %>%
  clean_names()
colnames(mydat2)
```
<br>


# Question 1: Kelp Growth & Size

## Distributions

### Response
```{r}
plotdat <- mydat1 %>%
  pivot_longer(cols=colnames(mydat1)[c(4:9)], names_to="variable")
ggplot(data=plotdat, aes(x=value)) + 
  geom_histogram() + 
  facet_wrap(~variable,ncol=2,scales="free")
```
<br>

It looks like all but mean holdfast diameter (`mapy_mean_hld`) may not be normal. 
```{r}
mydat1 %>%
  dplyr::select(mapy_mean_count,mapy_mean_density,mapy_max_fronds,mapy_mean_fronds,mapy_mean_hld,mapy_max_hld) %>%
  map(function(x) shapiro.test(as.numeric(x)))
```
<br>

Box-Cox suggested transformation:
```{r}
boxcox(lm(mapy_mean_density ~ urchin_abundance + site, data=mydat1),ylab="density log-L")
boxcox(lm(mapy_mean_fronds ~ urchin_abundance + site, data=mydat1),ylab="mean size log-L")
boxcox(lm(mapy_max_fronds ~ urchin_abundance + site, data=mydat1),ylab="max size log-L")
boxcox(lm(mapy_max_hld ~ urchin_abundance + site, data=mydat1),ylab="max holdfast size log-L")
```
<br>

Mean density and mean size should be log, whereas max size and is ~ 0.5 (square root)
```{r}
plotdat <- mydat1 %>%
  dplyr::select(mapy_mean_fronds,mapy_mean_density) %>%
  pivot_longer(cols=c(mapy_mean_fronds,mapy_mean_density))
ggplot(data=plotdat, aes(x=log(value))) + 
  geom_density() + 
  facet_wrap(~name,scales="free")
mydat1 %>%
  dplyr::select(mapy_mean_density,mapy_mean_fronds) %>%
  map(function(x) shapiro.test(as.numeric(log(x))))
```
<br>

Both measures of max size are ~ 0.5 (square root)
```{r}
plotdat <- mydat1 %>%
  dplyr::select(mapy_max_fronds,mapy_max_hld) %>%
  pivot_longer(cols=c(mapy_max_fronds,mapy_max_hld))
ggplot(data=plotdat, aes(x=sqrt(value))) + 
  geom_density() + 
  facet_wrap(~name,scales="free")
mydat1 %>%
  dplyr::select(mapy_max_fronds,mapy_max_hld) %>%
  map(function(x) shapiro.test(as.numeric(sqrt(x))))
```
<br>


# Question 2: Biodiversity

## Distributions

### Response
```{r}
plotdat <- dplyr::select(mydat2, -year, -month) %>%
  pivot_longer(cols=colnames(mydat2)[c(4:9)], names_to="variable")
ggplot(data=plotdat, aes(x=value)) + 
  geom_histogram() + 
  facet_wrap(~variable,ncol=3,scales="free")
```
<br>

It looks like all of these may not be normal. 
```{r}
mydat2 %>%
  dplyr::select(colnames(mydat2)[4:9]) %>%
  map(function(x) shapiro.test(as.numeric(x)))
```

Since algae and fish biomass / diversity follow approximately the same distribution, I'm going to just apply box-cox to fish responses and we can extend them to invertebrates.

Box-Cox suggested transformation:
```{r}
boxcox(lm(fish_diversity ~ kelp_total_biomass + site, data=mydat2),ylab="fish diversity log-L")
boxcox(lm(fish_total_biomass ~ kelp_total_biomass + site, data=mydat2),ylab="fish biomass log-L")
boxcox(lm(algae_diversity ~ kelp_total_biomass + site, data=mydat2),ylab="algae diversity log-L")
boxcox(lm(algae_total_biomass ~ kelp_total_biomass + site, data=mydat2),ylab="algae biomass log-L")
```

For fish / invert biomass, a log transformation? Invert biomass is still not normal...
```{r}
plotdat <- mydat2 %>%
  dplyr::select(fish_total_biomass,invert_total_biomass) %>%
  pivot_longer(cols=c(fish_total_biomass,invert_total_biomass))
ggplot(data=plotdat, aes(x=log(value))) + 
  geom_density() + 
  facet_wrap(~name,scales="free")
mydat2 %>%
  dplyr::select(fish_total_biomass,invert_total_biomass) %>%
  map(function(x) shapiro.test(as.numeric(log(x))))
```

For fish / invert diversity, $\lambda$ ~ -1. both are still not normal
```{r}
plotdat <- mydat2 %>%
  dplyr::select(fish_diversity,invert_diversity) %>%
  pivot_longer(cols=c(fish_diversity,invert_diversity))
ggplot(data=plotdat, aes(x=(1/value))) + 
  geom_density() + 
  facet_wrap(~name,scales="free")
mydat2 %>%
  dplyr::select(fish_diversity,invert_diversity) %>%
  map(function(x) shapiro.test(as.numeric((1/x))))
```


For algae diversity and biomass, a square root transformation.
```{r}
plotdat <- mydat2 %>%
  dplyr::select(algae_diversity,algae_total_biomass) %>%
  pivot_longer(cols=c(algae_diversity,algae_total_biomass))
ggplot(data=plotdat, aes(x=sqrt(value))) + 
  geom_density() + 
  facet_wrap(~name,scales="free")
mydat2 %>%
  dplyr::select(algae_diversity,algae_total_biomass) %>%
  map(function(x) shapiro.test(as.numeric(sqrt(x))))
```








