---
title: "504project"
author: "Tim Xi"
date: "2020/2/28"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
```{r}
library(tidyverse)
library(janitor)
library(vegan)
library(lubridate)
library(corrplot)
library(MASS)
library(here)
data.kelp<-read.csv(file='kelp_prediction_data_complete.csv')
data.waterchem<-read.csv(file='biomass_prediction_data_waterchem_complete.csv')
mydat2 <- read.csv(file='biomass_prediction_data_waterchem_bysite_complete.csv')                       
corrplot(cor(data.waterchem[,4:26],use="complete.obs"), type="upper")
```
```{r}
mydat2$kelp_total_biomass<-mydat2$kelp_total_biomass^(1/3)
mydat2$mean_pon<-log(mydat2$mean_pon)
mydat2$mean_tchl<-log(mydat2$mean_tchl)
mydat2$summer_mean_temp<-log(mydat2$summer_mean_temp)
mydat2$winter_mean_temp<-log(log(mydat2$winter_mean_temp))
```



```{r}
library(MASS)
boxcox(lm(fish_diversity~kelp_total_biomass+site, data=mydat2),ylab="fish diversity log-L")
boxcox(lm(fish_total_biomass ~kelp_total_biomass + site, data=mydat2),ylab="fish biomass log-L")
boxcox(lm(invert_diversity ~ kelp_total_biomass + site+summer_mean_temp+winter_mean_temp+mean_ammonia, data=mydat2),ylab="fish diversity log-L")
boxcox(lm(invert_total_biomass ~mean_tchl + summer_mean_temp + winter_mean_temp + 
    site + winter_mean_temp:site + summer_mean_temp:site, data=mydat2),ylab="fish biomass log-L")
boxcox(lm(algae_diversity ~ mean_ammonia+ winter_mean_temp, data=mydat2),ylab="algae diversity log-L")
boxcox(lm(algae_total_biomass ~ mean_no2_no3  + site, data=mydat2),ylab="algae biomass log-L")
```



preidctor transformed


#fish total biomass
```{r}
fit.initial<-lm(fish_total_biomass^(0.06)~mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+summer_mean_temp+winter_mean_temp+kelp_total_biomass+site, data=mydat2)
n<-dim(mydat2)[1]
scp<-list(lower=~1,upper=~mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+summer_mean_temp+winter_mean_temp+kelp_total_biomass+site+mean_poc:mean_pon:mean_tchl+mean_tchl:mean_pon+site:mean_tchl+mean_pon:site)
fit.final<-step(fit.initial, scope=scp, direction="backward",k=log(n))

```

```{r}
fit.null1<-lm(fish_total_biomass^(0.06)~1,data=mydat2)
fit.final.forward<-step(fit.null1,scope=scp,direction="forward",k=log(n))
summary(fit.final)
summary(fit.final.forward)
```


```{r}
par(mfrow=c(2,2))
plot(fit.final)
```

GLM
```{r}
fit1.glm<-glm(fish_total_biomass^(0.06)~mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+summer_mean_temp+winter_mean_temp+kelp_total_biomass+site, data=mydat2, family = Gamma)
summary(fit1.glm)
drop1(fit.final)
```

From above 
the final model we get is fish_total_biomass^(0.06) ~ mean_poc + log(mean_pon) + log(mean_tchl) + site



#diversity
```{r}
fit.initial2<-lm(fish_diversity^(-0.9)~mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+summer_mean_temp+winter_mean_temp+kelp_total_biomass+site,data=mydat2)
n<-dim(mydat2)[1]
scp2<-list(lower=~1,upper=~mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+summer_mean_temp+winter_mean_temp+kelp_total_biomass+site)
fit.final2<-step(fit.initial2, scope=scp2, direction="backward", k=log(n))
```

```{r}
fit.null2<-lm(fish_diversity^(-0.9)~1,data=mydat2)
fit.final2.forward<-step(fit.null2,scope=scp2,direction="forward",k=log(n))
summary(fit.final2)
```


```{r}
summary(glm(fish_diversity~kelp_total_biomass+mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+kelp_total_biomass:mean_poc+summer_mean_temp+winter_mean_temp+site, family="Gamma", data=mydat2))
```
The final model can be GLM with Gamma family 
fish_diversity~kelp_total_biomass^(1/3)+mean_ammonia+kelp_total_biomass:mean_poc


#invert_total_biomass
```{r}
fit.initial3<-lm(invert_total_biomass^(0.125)~mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+summer_mean_temp+winter_mean_temp+kelp_total_biomass+site,data=mydat2)
n<-dim(data.waterchem)[1]
scp3<-list(lower=~1,upper=~mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+summer_mean_temp+winter_mean_temp+kelp_total_biomass+site)
fit.final3<-step(fit.initial3,scope=scp3,direction="backward",k=log(n))
```

```{r}
fit.null3<-lm(invert_total_biomass^(0.125)~1,data=mydat2)
fit.final3.forward<-step(fit.null3,scope=scp3,direction="forward",k=log(n))
summary(fit.final3.forward)
summary(fit.final3)
```


so we take fit.final3, which is invert_total_biomass^(0.125) ~ log(mean_tchl) + site

```{r}
par(mfrow=c(2,2))
plot(fit.final3)
summary(fit.final3)
```

#invert_diversity
```{r}
fit.initial4<-lm(invert_diversity^(-2)~mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+summer_mean_temp+winter_mean_temp+kelp_total_biomass+site,data=mydat2)
n<-dim(mydat2)[1]
scp4<-list(lower=~1,upper=~mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+summer_mean_temp+winter_mean_temp+kelp_total_biomass+site)
fit.final4<-step(fit.initial4,scope=scp4,direction="backward",k=log(n))
```

```{r}
fit.null4<-lm(invert_diversity^(-2)~1,data=mydat2)
fit.final4.forward<-step(fit.null4,scope=scp4,direction="forward",k=log(n))
```

```{r}
summary(fit.final4)
summary(fit.final4.forward)
```

the final model is invert_diversity^(-2) ~ mean_ammonia + log(summer_mean_temp) + 
    kelp_total_biomass^(1/3) + site

```{r}
par(mfrow=c(2,2))
plot(fit.final4)
```
normality assumption seems to be satisfied but constant variance assumption seems to be violated.

#algae_total_biomass
```{r}
fit.initial5<-lm(algae_total_biomass^(0.45)~mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+summer_mean_temp+winter_mean_temp+kelp_total_biomass+site,data=mydat2)
n<-dim(mydat2)[1]
scp5<-list(lower=~1,upper=~mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+summer_mean_temp+winter_mean_temp+kelp_total_biomass+site)
fit.final5<-step(fit.initial5,scope=scp5,direction="backward",k=log(n))
```

```{r}
fit.null5<-lm(algae_total_biomass^(0.45)~1,data=mydat2)
fit.final5.forward<-step(fit.null5,scope=scp5,direction="forward",k=log(n))
summary(fit.final5)
```

```{r}
fit5.glm<-glm(algae_total_biomass^(0.45)~mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+summer_mean_temp+winter_mean_temp+kelp_total_biomass+site, data=mydat2, family = Gamma)
summary(fit5.glm)
```

we may take a GLM model with Gamma family
algae_total_biomass^(0.45) ~ mean_no2_no3 +site

#algae_diversity
```{r}
fit.initial6<-lm(algae_diversity^(0.25)~mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+summer_mean_temp+winter_mean_temp+kelp_total_biomass+site+winter_mean_temp:mean_ammonia,data=mydat2)
n<-dim(mydat2)[1]
scp6<-list(lower=~1,upper=~mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+summer_mean_temp+winter_mean_temp+kelp_total_biomass+site+winter_mean_temp:mean_ammonia)
fit.final6<-step(fit.initial6,scope=scp6,direction="backward",k=log(n))

```

```{r}
fit.null6<-lm(algae_diversity^(0.25)~1,data=mydat2)
fit.final6.forward<-step(fit.null6,scope=scp6,direction="forward",k=log(n))
summary(fit.final6)
```

```{r}
summary(glm(algae_diversity~mean_ammonia+mean_no2_no3+mean_po4+mean_poc+mean_pon+mean_tchl+summer_mean_temp+winter_mean_temp+kelp_total_biomass+site+winter_mean_temp:mean_ammonia, family="Gamma", data=mydat2))
```
the final model is algae_diversity^(0.5) ~ mean_ammonia + log(log(winter_mean_temp))
though it's not that good....

```{r}
par(mfrow=c(2,2))
plot(fit.final6)
```

